### 2023-11-26 thesis/sections/5-grounding.tex
background
methodology, diagrams

In a first step, bounding boxes are extracted from each scene.
Each bounding box is a square with a side length of 96 pixels around the center coordinates of each object.
By doing this large objects in the front of the scene fill the bounding box completely, while smaller objects in the back of the scene contain some space around them.
They also might contain noise in the form of parts of adjacent objects.
Preliminary experiments indicated however that a more complex method for extracting the bounding boxes, depending on spatial position in the scene to reduce the noise didn't improve the results.
Therefore, bounding boxes in this thesis rely on the former simpler approach.
Since, each bounding box will be passed through on of the feature extractors, described in section \ref{sec:feature-extractors} they need to be normalized first.
As described in \citep{He2016,Simonyan2015}, the bounding boxes are resized to 256 pixels for the shorter side, then cropped around the center to a square of 224 \times\ 224 pixels.
Finally, the RGB channels are normalized, by subtracting the means (0,485; 0,456; 0,406) and dividing the result by the standard deviation (0,229; 0,224; 0,225) for each channel respectively.
The array of bounding boxes is padded with matrix of zeros to the maximum possible number of objects present in a scene across the dataset.
For the 'Dale-2' dataset, this corresponds to a maximum of two bounding boxes, 5 bounding boxes for the 'Dale-5' dataset and 10 for the 'CLEVR color' dataset.
For each sample in the dataset the bounding boxes are shuffled.

@@

The bounding boxes are extracted and normalized in the same way as for the object identification task.
Since for the second setup, the images will also be passed through the feature extractors, they are normalized as well in like manner.
The referring expressions for the target object are generated using the incremental GRE-algorithm, described in section \ref{GRE}.
By this, the model needs to describe the target object with respect to the distractor objects.
There are some minor additions concerning the padding of the referring expression.
As before, the referring expression is padded to a number of three tokens, corresponding to the maximum of three attributes.
However, there are three different ways how the padding is applied.
First, the referring expression are, as usual in captioning tasks padded at the end with a specified padding token.
A problem could arise when the referring expression is not viewed as a natural language sentence, but as slots filled with tokens.
More specifically, following the GRE-algorithm, the last token in the referring expression is always the shape.
The second last token if existing describes the color, while the third last token if existing describes the size.
As soon as this sequence is padded at the end, these slots disappear.
A referring expression that only describes the shape, such as \emph{cube} will be padded to \emph{cube <pad> <pad>}, where the third last slot is filled up with the shape instead of the size.
Since this task is not focussing on producing natural language with a correct grammar, but focuses instead on extracting attributes, having a slot structure could help the model to express the extracted attributes correctly.
For this reason, the second method of padding the referring expression is prepending the referring expression with padding tokens.
By this, the positions of the slots are preserved and if not specified just filled with a padding token.
Each slot has always the same semantic value, e.g. the last slot always contains the shape of an object.
This can be done since the referring expressions are not free text, but instead the structure and the possible content is given by the dataset.
This method might help the model to learn the correct referring expressions.
% SD: Slots have the same semantic value, i.e. the last slot is the object type. We can do this because we know the structure of the descriptions in this dataset and because we hope this will help the learning. Also, we are thinking of generation based on slots.
% DK: (addressed; done)
The last variation concerns the order of producing each token.
When the referring expressions are prepended, the model would need first produce two padding tokens, before it finally can produce a much more meaningful token for the shape.
This could be difficult to learn for a model, as the longer a sequence of tokens is, the more information about the beginning of the sequence gets lost.
Even though a sequence of just three tokens may not be long enough for this factor to be a problem, we experimented to reverse the referring expression.
% SD: We... I sounds more like you are writing a lab report with casual language but here you want to be more formal.
% DK: (done)
Instead of producing for instance \emph{<pad> green sphere} as correct in English, the model would now need to produce \emph{sphere green <pad>}.
Notice that the padding token is again at the end of the generation, but the order of slots as well as the amount of information in the referring expression are still preserved.

@@

% TODO include figure

In the first setup, the \emph{bounding box RE generator}, the model receives the bounding boxes as input.
The first bounding box is always the target object, while the remaining bounding boxes are shuffled.
As in the previous experiment, each bounding box is passed through \emph{ResNet-avg} and projected afterwards to an image embedding dimension $e$ with a linear layer.
All thus encoded bounding boxes are concatenated and again compressed to the decoder output dimension $LSTM_o$ using another linear layer.
This representation of all objects serves as the initial hidden state of an LSTM, which generates the referring expression.
Tokens used in the LSTM are embedded with embedding dimension $LSTM_e$.
During training, teacher forcing is applied by using embeddings of the ground truth tokens as the input sequence for the LSTM, instead of the output of the LSTM.
The output of the LSTM is passed through a linear layer at each step to determine logits over the symbols of the vocabulary.
During testing, the LSTM is always forced to generate three tokens, with an embedded start-of-sequence token as first input to the LSTM.
Each token in the sequence is determined greedily, by selecting the highest logit in the output of each step in the LSTM.

\begin{figure}[ht]
    \centering
    \includegraphics[width=.5\linewidth]{figures/arch_caption_generator.png}
    \caption{Simplified architecture of the masked caption generator}
    \label{fig:caption_generator_architecture}
\end{figure}

@@

Using this approach, the model doesn't have any information about the target object and is therefore expected to produce referring expressions for a randomly selected object on the scene.
Therefore, it is extended in a second step with attention to the target object, the \emph{masked RE generator}.
% SD: You add attention to the target object.
% DK: (added; done)
A masked version of the image is created and passed to the model.
For this, a fixed size squared area of 96 \times\ 96 pixels around the center of the target object is filled in white, while the rest of the image is filled in black.
Both original and masked image are processed together with convolutional layers as described in section \ref{sec:image-processing}.
% SD: This is normally implemented as a filter of 1 and 0 corresponding to pixels. As such it then corresponds to attention, cf. paper with Mehdi.
% DK: (QUESTION)
The result is then again projected to the decoder output dimension $d$ by a linear layer and then used as the initial hidden state for the LSTM.
This should point the model towards which object to describe and discriminate from the distractors.

>>>>

In a first step, bounding boxes are extracted from each scene.
Each bounding box is a square with a side length of 96 pixels around the center coordinates of each object.
By doing this large objects in the front of the scene fill the bounding box completely, while smaller objects in the back of the scene contain some space around them.
They also might contain noise in the form of parts of adjacent objects.
Preliminary experiments indicated however that a more complex method for extracting the bounding boxes, depending on spatial position in the scene to reduce the noise didn't improve the results.
Therefore, bounding boxes in this thesis rely on the former simpler approach.
Since, each bounding box will be passed through on of the feature extractors, they are preprocessed as described in section \ref{sec:feature-extractors}.
The array of bounding boxes is padded with matrix of zeros to the maximum possible number of objects present in a scene across the dataset.
For the 'Dale-2' dataset, this corresponds to a maximum of two bounding boxes, 5 bounding boxes for the 'Dale-5' dataset and 10 for the 'CLEVR color' dataset.
For each sample in the dataset the bounding boxes are shuffled.

@@

The bounding boxes are extracted and normalized in the same way as for the object identification task.
Since for the second setup, the images will also be passed through the feature extractors, they are normalized as well in like manner.
The referring expressions for the target object are generated using the incremental GRE-algorithm, described in section \ref{sec:background:re}.
By this, the model needs to describe the target object with respect to the distractor objects.
There are some minor additions concerning the padding of the referring expression.
As before, the referring expression is padded to a number of three tokens, corresponding to the maximum of three attributes.
However, there are three different ways how the padding is applied.
First, the referring expression are, as usual in captioning tasks padded at the end with a specified padding token.
A problem could arise when the referring expression is not viewed as a natural language sentence, but as slots filled with tokens.
More specifically, following the GRE-algorithm, the last token in the referring expression is always the shape.
The second last token if existing describes the color, while the third last token if existing describes the size.
As soon as this sequence is padded at the end, these slots disappear.
A referring expression that only describes the shape, such as \emph{cube} will be padded to \emph{cube <pad> <pad>}, where the third last slot is filled up with the shape instead of the size.
Since this task is not focussing on producing natural language with a correct grammar, but focuses instead on extracting attributes, having a slot structure could help the model to express the extracted attributes correctly.
For this reason, the second method of padding the referring expression is prepending the referring expression with padding tokens.
By this, the positions of the slots are preserved and if not specified just filled with a padding token.
Each slot has always the same semantic value, e.g. the last slot always contains the shape of an object.
This can be done since the referring expressions are not free text, but instead the structure and the possible content is given by the dataset.
This method might help the model to learn the correct referring expressions.
% SD: Slots have the same semantic value, i.e. the last slot is the object type. We can do this because we know the structure of the descriptions in this dataset and because we hope this will help the learning. Also, we are thinking of generation based on slots.
% DK: (addressed; done)
The last variation concerns the order of producing each token.
When the referring expressions are prepended, the model would need first produce two padding tokens, before it finally can produce a much more meaningful token for the shape.
This could be difficult to learn for a model, as the longer a sequence of tokens is, the more information about the beginning of the sequence gets lost.
Even though a sequence of just three tokens may not be long enough for this factor to be a problem, we experimented to reverse the referring expression.
% SD: We... I sounds more like you are writing a lab report with casual language but here you want to be more formal.
% DK: (done)
Instead of producing for instance \emph{<pad> green sphere} as correct in English, the model would now need to produce \emph{sphere green <pad>}.
Notice that the padding token is again at the end of the generation, but the order of slots as well as the amount of information in the referring expression are still preserved.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=.5\linewidth]{figures/arch_bb_caption_generator.png}
    \caption{Simplified architecture of the bounding box RE generator}
    \label{fig:bb_re_generator_architecture}
\end{figure}

In the first setup, the \emph{bounding box RE generator}, the model receives the bounding boxes as input (see Figure \ref{fig:bb_re_generator_architecture}).
The first bounding box is always the target object, while the remaining bounding boxes are shuffled.
As in the previous experiment, each bounding box is passed through \emph{ResNet-avg} and projected afterwards to an image embedding dimension $e$ with a linear layer.
All thus encoded bounding boxes are concatenated and again compressed to the decoder output dimension $LSTM_o$ using another linear layer.
This representation of all objects serves as the initial hidden state of an LSTM, which generates the referring expression.
Tokens used in the LSTM are embedded with embedding dimension $LSTM_e$.
During training, teacher forcing is applied by using embeddings of the ground truth tokens as the input sequence for the LSTM, instead of the output of the LSTM.
The output of the LSTM is passed through a linear layer at each step to determine logits over the symbols of the vocabulary.
During testing, the LSTM is always forced to generate three tokens, with an embedded start-of-sequence token as first input to the LSTM.
Each token in the sequence is determined greedily, by selecting the highest logit in the output of each step in the LSTM.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=.5\linewidth]{figures/arch_caption_generator.png}
    \caption{Simplified architecture of the masked RE generator}
    \label{fig:masked_re_generator_architecture}
\end{figure}

Using this approach, the model doesn't have any information about the target object and is therefore expected to produce referring expressions for a randomly selected object on the scene.
Therefore, it is extended in a second step with attention to the target object, the \emph{masked RE generator} (see Figure \ref{fig:masked_re_generator_architecture}).
% SD: You add attention to the target object.
% DK: (added; done)
A masked version of the image is created and passed to the model.
For this, a fixed size squared area of 96 \times\ 96 pixels around the center of the target object is filled in white, while the rest of the image is filled in black.
Both original and masked image are processed together with convolutional layers as described in section \ref{sec:image-processing}.
% SD: This is normally implemented as a filter of 1 and 0 corresponding to pixels. As such it then corresponds to attention, cf. paper with Mehdi.
% DK: (done)
The result is then again projected to the decoder output dimension $d$ by a linear layer and then used as the initial hidden state for the LSTM.
This should point the model towards which object to describe and discriminate from the distractors.
